What We Eval in the Shadows

### 2021-04-17 paper/main.tex
sec 5

\medskip\noindent\EE{$min(e)=$~\c{<-}} This form includes both assignment
operators, the direct assignment {\tt <-}, and assignment to the parent
environment {\tt <\,\!<-}, the \c{assign} function and the \c{\$} form. Assignments occur in 5\% of the
sites. They represent the most obvious source of side effects. The median number
of operations is \packageMinimizedmedianoperationsiRnd; the minimum is 3.

\medskip\noindent\EE{$min(e)=\c{F()}$} This form captures simple function calls,
\eg~\c{f(2)}, with neither variables nor assignments. More specifically, it
allows for variables in the function position but not in arguments. Usually,
looking up function names does not trigger computation, but that is not a given
if the function is returned by a promise or if the function name is shadowed by
a promise containing a value, then the computation will occur. This form occurs in
6\% of sites and typically does not perform much work in the interpreter.

@@

\paragraph{Consistency} It is interesting to consider how many different
forms any given site sees. The more forms, the harder it will be to characterize
the program's behavior at that site. Luckily, \packageNbOneMinimizedPercent
of sites only see a single form.  In other words, a JIT compiler could correctly predict the \emph{shape} of an expression by looking just at the first call in \packageNbOneMinimizedPercent of the sites.  Few sites are highly polymorphic (8 or more different forms). These include the pipe operator of the \c{magrittr} package, which is used to compose functions.

\paragraph{Discussion} The variety of uses of \eval is evidenced by the number
of different forms observed in CRAN. In comparison, JavaScript \eval usage was more straightforward and more predictable as reported by \citet{oopsla12b}, with 98.7\% of the sites with only one form. Nevertheless, simple forms dominate in R also, and there are many cases where \eval could be replaced by less powerful constructs.

@@

\subsection{Side effects in \eval}

@@

From the \packageNbrunsRnd programs, we capture \SEAllRnd side effects from
\SEAllCallsRnd \eval calls (\SEUserCallsToAllRatio) in \SEAllSites sites
(\SEUserSitesToAllRatio). Again, the challenge is to remove the accidental
side effects caused by the R virtual machine implementation that are not
related to the user code. For example, the \c{.Random.seed} variable, which
contains the state of the random number generator, is saved and restored from
and to the global environment every time the user calls one of the base
routines that use random numbers.
%
% \begin{wrapfigure}{r}{5cm} \hspace*{-20mm} \centering
% \includegraphics[width=90mm]{se-types.pdf} \caption{Distribution of R types
% in target environments for side effects.} \label{fig:se-types}
% \end{wrapfigure}
%
Removing them leaves us with \SEUserRnd side effects (\SEUserRatio) from
\SEUserCallsRnd \eval calls (\SEUserCallsRatio) in \SEUserSites sites
(\SEUserSitesRatio). These sites are part of \SEUserFunctions R functions
(\SEUserFunctionsToAllRatio of all functions doing \eval) in \SEUserPackages
packages. From these functions, \SEFunsNighty are responsible for 90\% of side
effects. Half of the side-effects come just from three functions:
\c{plyr::allocate\_column} (allocates space for a new data frame column),
\c{withr::execute\_handlers} (executes deferred expressions), and
\c{foreach::doSEQ} (executes an expression on each element in a collection,
possibly in parallel). Most of the \eval sites (\SESitesInEnvirRatio) do side
effects in the environment specified by the \c{envir} parameter (\cf
Section~\ref{sec:eval-in-r}), \SESitesNotInEnvirRatio modifies other
environments, and finally, \SESitesBothEnvirRatio does both.
Table~\ref{tab:se-env} shows the class of the environment where \eval side
effects happen (\emph{target environment}). It follows the classification
described in Section~\ref{sec:env} with two extra classes: \emph{Object},
which represents a side-effect that happens in S4 or R6 object environments, and
\emph{Local} that corresponds to a side effect in local variables of the \eval
caller (offset 0 in Section~\ref{sec:env}).  The table shows data for both \eval sites and functions. For a
function to have a given target environment class, all of its \eval sites that
do side effects must have the same target environment class The majority of the
\eval sites (\SESitesInOneClass) do all side effects consistently in one
environment class. The same happens at the function level. Almost half of the
sites (over a third of the functions) do side effects in either \emph{Local} or
\emph{Object} environments. This gives a ray of hope for the compiler. Even
though it is possible to do anything anywhere, the data suggest that most
side-effects are sane.

@@

>>>>

\medskip\noindent\EE{$min(e)=$~\c{<-}} This form includes both assignment
operators, the direct assignment {\tt <-}, and assignment to the parent
environment {\tt <\,\!<-}, the \c{assign} function and the \c{\$} form.
Assignments occur in 5\% of the sites. They represent the most obvious source of
side effects. The median count is \packageMinimizedmedianoperationsiRnd (min=3).

\medskip\noindent\EE{$min(e)=\c{F()}$} This form captures simple function calls,
\eg~\c{f(2)}, with neither variables nor assignments. More specifically, it
allows for variables in the function position but not in arguments. Usually,
looking up function names does not trigger computation, but that is not a given
if the function is returned by a promise or if the function name is shadowed by
a promise containing a value, then the computation will occur. This form occurs in
6\% of sites and typically does not perform much work.

@@

\paragraph{Consistency} It is interesting how many different
forms any given site sees. The more forms, the harder it is to characterize
behavior at that site. Luckily, \packageNbOneMinimizedPercent of sites only see
a single form. In other words, a compiler could correctly predict the
\emph{shape} of an expression by looking just at the first call in
\packageNbOneMinimizedPercent of the sites. Few sites are highly polymorphic (8
or more different forms). These include the pipe operator of the \c{magrittr}
package, used to compose functions.

\paragraph{Discussion} The variety of uses of \eval is evidenced by the number
of different forms. In comparison, JavaScript \eval usage was more
straightforward and more predictable as reported by \citet{oopsla12b}, with
98.7\% of the sites with only one form. Nevertheless, simple forms dominate in R
also, and there are many cases where \eval could be replaced by less powerful
constructs.

@@

\subsection{The effects of \eval}

@@

From the \packageNbrunsRnd programs, we capture \SEAllRnd side effects from
\SEAllCallsRnd \eval calls (\SEUserCallsToAllRatio) in \SEAllSites sites
(\SEUserSitesToAllRatio). Again, the challenge is to remove the accidental
side effects caused by the R virtual machine implementation that are not
related to the user code. For example, the \c{.Random.seed} variable, which
contains the state of the random number generator, is saved and restored from
and to the global environment every time the user calls one of the base
routines that use random numbers.
%
% \begin{wrapfigure}{r}{5cm} \hspace*{-20mm} \centering
% \includegraphics[width=90mm]{se-types.pdf} \caption{Distribution of R types
% in target environments for side effects.} \label{fig:se-types}
% \end{wrapfigure}
%
Removing them leaves us with \SEUserRnd side effects (\SEUserRatio) from
\SEUserCallsRnd \eval calls (\SEUserCallsRatio) in \SEUserSites sites
(\SEUserSitesRatio). These sites are part of \SEUserFunctions R functions
(\SEUserFunctionsToAllRatio of all functions doing \eval) in \SEUserPackages
packages. From these functions, \SEFunsNighty are responsible for 90\% of side
effects. Half of the side-effects come just from three functions:
\c{plyr::allocate\_column} (allocates space for a new data frame column),
\c{withr::execute\_handlers} (executes deferred expressions), and
\c{foreach::doSEQ} (executes an expression on each element in a collection,
possibly in parallel). Most of the \eval sites (\SESitesInEnvirRatio) do side
effects in the environment specified by the \c{envir} parameter (\cf
Section~\ref{sec:eval-in-r}), \SESitesNotInEnvirRatio modifies other
environments, and finally, \SESitesBothEnvirRatio does both.

@@

Table~\ref{tab:se-env} shows the class of the environment where \eval side
effects happen (\emph{target environment}). It follows the classification
described in Section~\ref{sec:env} with two extra classes: \emph{Object}, which
represents a side-effect that happens in S4 or R6 object environments, and
\emph{Local} that corresponds to a side effect in local variables of the \eval
caller (offset 0 in Section~\ref{sec:env}). The table shows data for both \eval
sites and functions. For a function to have a given target environment class,
all of its \eval sites that do side effects must have the same target
environment class The majority of the \eval sites (\SESitesInOneClass) do all
side effects consistently in one environment class. The same happens at the
function level. Almost half of the sites (over a third of the functions) do side
effects in either \emph{Local} or \emph{Object} environments. This gives a ray
of hope for the compiler. Even though it is possible to do anything anywhere,
the data suggest that most side-effects are sane.
